Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-17 paper2.qmd
section:4,5; process:write,formulas

<!-- 1. The structural approach to causal inference -->
The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data. The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments. Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests. *Identification analysis* helps researchers to determine whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023]. Here, *estimands* represent the specific quantities researchers aim to determine [@Everitt_et_al_2010]. *Estimators* denote the methods or functions that transform data into an estimate, while *estimates* are the numerical values approximating the estimand [@Neal_2020; @Everitt_et_al_2010].

@@

<!-- 1. What are we going to do? -->
Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023] and Item Response Theory (IRT) [@Fox_2010; @vanderLinden_et_al_2017_I], the conceptual-population model integrates two *hierarchical structural components* to examine how different teaching methods influence students' writing ability. Each structural component defines how observed or latent variables affect the primary latent variable of interest [@Everitt_et_al_2010]. Their hierarchical nature enables researchers to test hypotheses considering the hierarchical structure of stimuli (see @sec-theory-issue1b) and the uncertainties in trait estimation (see @sec-theory-issue2).

@@

<!-- 2 Explaining what we did -->
The top branch of [DAG -@fig-cj09_dag] illustrates the first component, where *relevant* [^2] student-related variables $(X_{I})$, such as teaching method, and students' idiosyncratic errors $e_{I}$ causally influence the latent variable representing students' writing-quality trait $(T_{I})$. The error term $e_{I}$ captures variations in students' traits unexplained by $X_{I}$. Here, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$. Additionally, this branch shows how $T_{I}$, along with *relevant* [^3] text-related variables $(X_{IA})$ (e.g., text length), and texts' idiosyncratic errors $e_{IA}$ causally influence the texts' written-quality trait $(T_{IA})$, the first primary latent variable of interest. The error term $e_{IA}$ captures variations in the texts' traits that remain unexplained by $T_{I}$ or $X_{IA}$. Here, $X_{IA}$ is an observed matrix with dimensions $n_{I} \cdot n_{A}$ rows and $q_{IA}$ columns (variables), while $e_{IA}$ and $T_{IA}$ are latent matrices with $n_{I}$ rows and $n_{A}$ columns.

@@

Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^4] judge-related variables $(Z_{J})$, such as judgment expertise, and judges' idiosyncratic errors $e_{J}$ causally influence the latent variable representing judges' bias $(B_{J})$. The error $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$. Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$. Furthermore, the branch shows how $B_{J}$, along with *relevant* [^5] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the judges' biases associated with each text $B_{JK}$, the second primary latent variable of interest. The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$. Here, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent latent matrices with $n_{J}$ rows and $n_{K}$ columns

@@

<!-- 1. What is the more realistic scenario: the sample mechanism -->
The sample-comparison model presents a more realistic scenario than the conceptual-population model. In @sec-theory-theoretical_SC1, it explicitly assumes researchers work with a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population. Furthermore, in @sec-theory-theoretical_SC2, the model assumes that judges do not perform *all repeated judgments* within the data sample. Instead, they conduct a sufficient number of stimuli comparisons, $n_{C}$, to ensure an accurate estimation of the proportion $P(B>A)$, as proposed by @Thurstone_1927b.

@@

<!-- 1. start by defining the comparison vector  -->
As in the previous section, we begin defining the *comparison mechanism* using the binary vector variable $C$ to facilitate the interpretation of the sample-comparison model (see equation ([-@eq-mat27])). The vector contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(i,a,h,b,j,k)}$ being a binary value indicating the presence or absence of data rows in $R$, a definition similar to that of $s_{(i,a,h,b,j,k)}$ in equation ([-@eq-mat26]).
$$
C = \begin{bmatrix}
c_{(1,1,1,2,1,1)} \\
\vdots \\
c_{(1,1,1,2,1,n_{K})} \\
\vdots \\
c_{(i,a,h,b,j,k)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
$$ {#eq-mat27}

<!-- 2. What is the effect of the comparison mechanism -->
The [DAG -@fig-cj14_dag] also integrates the *comparison mechanism* $C$ into the conceptual-population model, ultimately deriving the outcome $O_{R}$. It shows the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access to $O^{sc}_{R}$ due to the comparison mechanism. Moreover, the DAG depicts $C$ as a causal factor influencing $O_{R}$, enclosing it in a square to indicate that it is a conditioned variable. Specifically, $C$ determines *which repeated judgments the subset of judges makes for the subset of stimuli produced by the sampled individuals*. In essence, $C$ reflects the assumption that judges *do not* perform all possible repeated judgments for the selected stimuli. Instead, they carry out a "sufficient number of comparisons" $n_{C}$ to allow for accurate estimation of the proportion $P(B>A)$ for each stimulus pair [@Thurstone_1927b, pp. 267].

@@

<!-- 4. segway for next section -->
In summary, the [SCM -@fig-cj15_scm] and [DAG -@fig-cj15_dag] extend Thurstone's general form to address several limitations of the BTL model. These extensions account for judge biases (see @sec-theory-issue1b), reflect the hierarchical structure of stimuli (see @sec-theory-issue1b), and incorporate measurement error in trait estimation and hypothesis testing (see @sec-theory-issue2). However, they do not resolve concerns about the assumption of equal dispersions among stimuli discussed in @sec-theory-issue1a. Because this issue involves statistical assumptions about the distribution of the discriminal process, it requires a statistical model, which we develop in the next section.

@@

<!-- 1. How to derive an stat model from an SCM? -->
With the structure in [SCM -@fig-cj15_scm], we can derive a statistical model that addresses violations of the equal dispersion assumption in CJ experiments (see @sec-theory-issue1a). This derivation is possible because a fully specified SCM encodes functional and probabilistic information, which we can replace with appropriate functions and probabilistic assumptions [@Pearl_et_al_2016]. Specifically, [SCM -@fig-cj15_scm] allows us to express the joint distribution of our complex CJ system as a product of simpler conditional probability distributions (CPDs)[^6], as shown in equation ([-@eq-mat51]). For clarity, we treat expressions such as $Y := f_{Y}(X)$, $P(Y \mid X)$, and $Y \sim f(Y \mid X)$ as equivalent. Here, $P(Y \mid X)$ and $f(Y \mid X)$ represent the CPD of $Y$ given $X$.
$$
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
$$ {#eq-mat51}

@@

<!-- 2. assumptions about the outcome, and the sample and comparison mechanism -->
Each CPD in equation ([-@eq-mat51]) rests on specific assumptions, which we outline in the [SCM -@fig-cj16_scm], the probabilistic model [-@fig-cj16_scm], and the statistical model [-@fig-cj16_scm]. First, we assume that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes. Furthermore, following the conventions of Generalized Linear Latent and Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$. Note that while the probabilistic model [-@fig-cj16_prob] includes the probability distributions for the sampling and comparison mechanisms, $P(S)$ and $P(C)$, we omit these from the statistical model [-@fig-cj16_stat]. We do so because, although these distributions contribute to the joint distribution of the data, they do not affect the relationship between $O_{R}$ and $D_{R}$. This lack of influence is due to the random selection designs that govern $S$ and $C$[^8].

@@

[^8]: Randomization ensures that data—and, by extension, an estimator—satisfies several key identification properties, such as common support, no interference, and consistency. The most critical property, however, is the elimination of confounding. *Confounding* occurs when an external variable, such as $X_{I}$, simultaneously influences both the outcome (e.g., $O_{R}$) and a variable of interest (e.g., $S$), resulting in spurious associations between the two [@Everitt_et_al_2010]. Randomization addresses this issue by effectively decoupling the association between the variable of interest and any other variable, except for the outcome itself. For a more detailed discussion on the benefits of randomization, see @Pearl_2009, @Morgan_et_al_2014, @Neal_2020, and @Hernan_et_al_2020.

@@

::: {#fig-cj16_prob fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  & P( O_{R} \mid D_{R}, S, C ) \cdot P(S) \cdot P(C) \\
  & P( D_{R} \mid T_{IA}, B_{JK} ) \\
  & P( T_{IA} \mid T_{I}, X_{IA}, e_{IA} ) \\
  & P( T_{I} \mid X_{I}, e_{I} ) \\
  & P( B_{JK} \mid B_{J}, Z_{JK}, e_{JK} ) \\
  & P( B_{J} \mid Z_{J}, e_{J} ) \\
  & P( e_{I} ) \\
  & P( e_{J} ) \\
  & P( e_{IA} ) \\  
  & P( e_{JK} )
\end{aligned}
$$

@@

Sixth, we define $B_{J}$ as a linear combination of the effects of relevant judge-level variables $\beta_{ZJ}Z_{J}$ (e.g., judgment expertise) and judge-specific idiosyncratic errors $e_{J}$. Each element of $e_{J}$ is likewise assumed to be $iid$ Normal with mean zero and a standard deviation $s_{ZJ}$ that depends on the groups to which each judge belongs. For instance, if $Z_{J}$ is a variable that can take on three values representing three groups of judges with varying expertise, the model imposes the constraint $\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1$ to fix the scale of the latent trait [@Depaoli_2021].

@@

Finally, to transform [model -@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data, we use *Bayesian inference methods*. The **Declarations** section of this document includes a link to the model, along with an alternative specification that assumes equal discriminal dispersions. We implemented both versions of the model using `Stan` [version $2.26.1$, @Stan_2020]. Although the implementation was successful, we do not report the results here, as they fall outside the scope of this document.

@@

<!-- 2. What are the benefits of Bayesian inference?-->
<!-- Bayesian inference procedures offer three key advantages. First, they are well-suited to handling complex and overparameterized models, enabling researchers to estimate models where the number of parameters exceeds the number of observations for estimation [@Baker_1998; @Kim_et_al_1999]. Second, they allow researchers to incorporate prior information, which helps constrain parameters within specified bounds. This capability addresses challenges such as non-convergence or improper parameter estimation that often arise in complex models when analyzed with frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011]. Finally, Bayesian methods are particularly effective at drawing inferences from small sample sizes, where relying on the asymptotic properties of frequentist approaches may not be justified [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014]. -->

>>>>

<!-- 1. The structural approach to causal inference -->
The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data. The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments. Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests. *Identification analysis* determines whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023]. Here, *estimands* represent the specific quantities researchers aim to determine [@Everitt_et_al_2010]. *Estimators* denote the methods or functions that transform data into an estimate, while *estimates* are the numerical values approximating the estimand [@Neal_2020; @Everitt_et_al_2010].

@@

<!-- 1. What are we going to do? -->
Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023] and Item Response Theory (IRT) [@Fox_2010; @vanderLinden_et_al_2017_I], the conceptual-population model integrates two *hierarchical structural components* to examine how different teaching methods influence students' writing ability. Each structural component defines how observed or latent variables affect the primary latent variable of interest [@Everitt_et_al_2010]. The model's hierarchical design allows researchers to formulate and test hypotheses that account for both the nested structure of stimuli and the uncertainties inherent in trait estimation (see @sec-theory-issue1b and @sec-theory-issue2 for further discussion of these considerations).

@@

<!-- 2 Explaining what we did -->
The top branch of [DAG -@fig-cj09_dag] illustrates the first component, where *relevant* [^2] student-related variables $(X_{I})$, such as teaching method, and students' idiosyncratic errors $e_{I}$ causally influence the latent variable representing students' writing-quality trait $(T_{I})$. The error term $e_{I}$ captures variations in students' traits unexplained by $X_{I}$. Here, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ independent columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$. Additionally, this branch shows how $T_{I}$, along with *relevant* [^3] text-related variables $(X_{IA})$ (e.g., text length), and texts' idiosyncratic errors $e_{IA}$ causally influence the texts' written-quality trait $(T_{IA})$, the first primary latent variable of interest. The error term $e_{IA}$ captures variations in the texts' traits that remain unexplained by $T_{I}$ or $X_{IA}$. Here, $X_{IA}$ is an observed matrix with dimensions $n_{I} \cdot n_{A}$ rows and $q_{IA}$ independent columns (variables), while $e_{IA}$ and $T_{IA}$ are latent matrices with $n_{I}$ rows and $n_{A}$ columns.

@@

Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^4] judge-related variables $(Z_{J})$, such as judgment expertise, and judges' idiosyncratic errors $e_{J}$ causally influence the latent variable representing judges' bias $(B_{J})$. The error $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$. Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ independent columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$. Furthermore, the branch shows how $B_{J}$, along with *relevant* [^5] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the judges' biases associated with each text $B_{JK}$, the second primary latent variable of interest. The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$. Here, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ independent columns (variables), while $e_{JK}$ and $B_{JK}$ are latent latent matrices with $n_{J}$ rows and $n_{K}$ columns

@@

<!-- 1. What is the more realistic scenario: the sample mechanism -->
The sample-comparison model presents a more realistic scenario than the conceptual-population model. First, in @sec-theory-theoretical_SC1, it explicitly assumes researchers work with a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population. Second, in @sec-theory-theoretical_SC2, the model assumes that judges do not perform *all repeated judgments* within the data sample. Instead, they conduct a sufficient number of stimuli comparisons, $n_{C}$, to ensure an accurate estimation of the proportion $P(B>A)$, as proposed by @Thurstone_1927b.

@@

<!-- 1. start by defining the comparison vector  -->
As in the previous section, we begin defining the *comparison mechanism* using the binary vector variable $C$ to facilitate the interpretation of the sample-comparison model. Equation ([-@eq-mat27]) shows that $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(i,a,h,b,j,k)}$ being a binary value indicating the presence or absence of data rows in $R$, a definition similar to that of $s_{(i,a,h,b,j,k)}$ in equation ([-@eq-mat26]).
$$
C = \begin{bmatrix}
c_{(1,1,1,2,1,1)} \\
\vdots \\
c_{(1,1,1,2,1,n_{K})} \\
\vdots \\
c_{(i,a,h,b,j,k)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
$$ {#eq-mat27}

<!-- 2. What is the effect of the comparison mechanism -->
The [DAG -@fig-cj14_dag] also integrates the *comparison mechanism* $C$ into the conceptual-population model, ultimately deriving the outcome $O_{R}$. It shows the sample-comparison outcome as unobserved, highlighting that researchers cannot directly access to $O^{sc}_{R}$ due to the comparison mechanism. Moreover, the DAG depicts $C$ as a causal factor influencing $O_{R}$, enclosing it in a square to indicate that it is a conditioned variable. Specifically, $C$ determines *which repeated judgments the subset of judges makes for the subset of stimuli produced by the sampled individuals*. In essence, $C$ reflects the assumption that judges *do not* perform all possible repeated judgments for the selected stimuli. Instead, they carry out a "sufficient number of comparisons" $n_{C}$ to allow for accurate estimation of the proportion $P(B>A)$ for each stimulus pair [@Thurstone_1927b, pp. 267].

@@

<!-- 4. segway for next section -->
In summary, the [SCM -@fig-cj15_scm] and [DAG -@fig-cj15_dag] extend Thurstone's general form to address several limitations of the BTL model. These extensions account for judge biases (see @sec-theory-theoretical_P1), reflect the hierarchical structure of stimuli (see @sec-theory-theoretical_P3), incorporate measurement error in trait estimation and hypothesis testing (also see @sec-theory-theoretical_P3), and clarify the role of the sample and comparison mechanisms in CJ experiments (see @sec-theory-theoretical_SC). However, they do not resolve concerns about the assumption of equal dispersions among stimuli discussed in @sec-theory-issue1a. Because this concern relates to the statistical assumption underlying the distribution of the discriminal process, we develop a formal statistical model to address it in the next section.

@@

<!-- 1. How to derive an stat model from an SCM? -->
With the structure in [SCM -@fig-cj15_scm], we can derive a statistical model that addresses violations of the equal dispersion assumption in CJ experiments (see @sec-theory-issue1a). This derivation is possible because a fully specified SCM encodes functional and probabilistic information, which we can replace with suitable functions and probabilistic assumptions [@Pearl_et_al_2016]. Specifically, [SCM -@fig-cj15_scm] allows us to express the joint distribution of our complex CJ system as a product of simpler conditional probability distributions (CPDs)[^6], as shown in equation ([-@eq-mat51]). For clarity, we treat expressions such as $Y := f_{Y}(X)$, $P(Y \mid X)$, and $Y \sim f(Y \mid X)$ as equivalent. Here, $P(Y \mid X)$ and $f(Y \mid X)$ represent the CPD of $Y$ given $X$.
$$
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
$$ {#eq-mat51}

@@

<!-- 2. assumptions about the outcome, and the sample and comparison mechanism -->
Each CPD in equation ([-@eq-mat51]) rests on specific assumptions, which we outline in the [SCM -@fig-cj16_scm], the probabilistic model [-@fig-cj16_scm], and the statistical model [-@fig-cj16_scm]. First, we assume that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes. Furthermore, following the conventions of Generalized Linear Latent and Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.

@@

::: {#fig-cj16_prob fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  & P( O_{R} \mid D_{R}, S, C ) \\
  & P( D_{R} \mid T_{IA}, B_{JK} ) \\
  & P( T_{IA} \mid T_{I}, X_{IA}, e_{IA} ) \\
  & P( T_{I} \mid X_{I}, e_{I} ) \\
  & P( B_{JK} \mid B_{J}, Z_{JK}, e_{JK} ) \\
  & P( B_{J} \mid Z_{J}, e_{J} ) \\
  & P( e_{I} ) \\
  & P( e_{J} ) \\
  & P( e_{IA} ) \\  
  & P( e_{JK} )
\end{aligned}
$$

Note that while the joint distribution in Equation ([-@eq-mat51]) includes the probability distributions of the sampling and comparison mechanisms, $P(S)$ and $P(C)$, as well as those of the predictor variables--$P(X_{IA})$, $P(X_{I})$, $P(Z_{JK})$, and $P(Z_{J})$--we omit these probabilities from the models presented in @fig-cj16. We make this omission because, although these distributions contribute to the overall joint distribution of the data, they do not influence the relationship between $O_{R}$ and $D_{R}$. This independence stems from the random selection designs that govern $S$, $C$, $X_{IA}$, $X_{I}$, $Z_{JK}$, and $Z_{J}$[^8].

[^8]: Randomization ensures that data—and, by extension, an estimator—satisfies several key identification properties, such as common support, no interference, and consistency. The most critical property, however, is the elimination of confounding. *Confounding* occurs when an external variable, such as $X_{I}$, simultaneously influences both the outcome (e.g., $O_{R}$) and a variable of interest (e.g., $X_{I}$ or $S$), resulting in spurious associations between the two [@Everitt_et_al_2010]. Randomization addresses this issue by effectively decoupling the association between the variable of interest and any other variable, except for the outcome itself. For a more detailed discussion on the benefits of randomization, see @Pearl_2009, @Morgan_et_al_2014, @Neal_2020, and @Hernan_et_al_2020.

@@

@@

Sixth, we define $B_{J}$ as a linear combination of the effects of relevant judge-level variables $\beta_{ZJ}Z_{J}$ (e.g., the effects of judgment expertise on the quality assessment) and judge-specific idiosyncratic errors $e_{J}$. Each element of $e_{J}$ is likewise assumed to be $iid$ Normal with mean zero and a standard deviation $s_{ZJ}$ that depends on the groups to which each judge belongs. For instance, if $Z_{J}$ is a variable that can take on three values representing three groups of judges with varying expertise, the model imposes the constraint $\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1$ to fix the scale of the latent trait [@Depaoli_2021].

@@

Finally, using *Bayesian inference methods* to transform the [model -@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data. Bayesian inference offers three key advantages. First, it is well-suited for estimating complex and overparameterized models, where the number of parameters may exceed the number of observations [@Baker_1998; @Kim_et_al_1999]. Second, it allows researchers to incorporate prior information, which constrains parameter estimates within plausible bounds. This feature mitigates common estimation issues, such as non-convergence or improper solutions, often encountered in complex models under frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011]. Third, Bayesian inference is particularly effective for drawing conclusions from small samples, where the asymptotic properties underlying frequentist methods are less reliable [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014]. The **Declarations** section of this document provides a link to the model code, along with an alternative specification that assumes equal discriminal dispersions. We implemented both versions of the model using `Stan` [@Stan_2020, version 2.26.1]. Although the implementations were successful, we do not report the results here, as they fall outside the scope of this document.

@@
